Do syntactic and semantic similarity lead to interference effects?\\ Evidence from self-paced reading and event-related potentials using German

### 2024-11-26 paper/revision2/elsarticle-template-harv.tex
Updated via the Open Science Framework

\textcolor{blue}{Crucially, the reading times difference starting at the distractor} persisted in the following regions. The distractor and the immediately following regions differed between conditions, \textcolor{blue}{therefore we did not analyze their reading times. The reading times difference starting at the distractor and persisting almost until the end of the sentence is problematic for the analysis and interpretation of reading times in the later regions, i.e., the critical region. Any effects that might be present there cannot be attributed clearly to the processing of the respective region. However, for comparability with previous work, we briefly report analyses of the reading times of the later regions, i.e., the pre-critical, critical and spill-over region.}

@@

\paragraph{Hypothesis testing using Bayes factors}
\textcolor{blue}{To complete the analysis of the effects in the critical and surrounding regions -- although they are confounded by the long-lasting reading time differences starting at the distractor, we here present a Bayes factor analysis using all the data from Experiments 1a and 1b (see Figure \ref{fig:spr_bfs}). There was} very strong to extreme evidence for the semantic interference effect in the pre-critical and spill-over region (BF$_{10}$ between 36 and 231). In the critical region, the evidence for semantic interference was weak \textcolor{blue}{at best} (Normal(0, 0.01): BF$_{10}$ = 3.5, Normal(0, 0.05): BF$_{10}$ = 1.3, Normal(0, 0.1): BF$_{10}$ = 0.7). There was evidence against an effect of syntactic interference under almost all priors in all regions (BF$_{10}$ $<$ 0.5). Only in the spill-over region when small effects between -8 and 8 ms were assumed a priori, there was anecdotal evidence for syntactic interference (BF$_{10}$ = 2.8). There was overall evidence against the interaction of syntactic and semantic interference (BF$_{10}$ $<$ 1). Under the narrowest prior in the critical region, the Bayes factor provided no evidence either way (neither for nor against the interaction (BF$_{10}$ = 1.4)).

@@

\subsection*{Discussion}
\Copy{SPR_discussion_revised}{We presented the to-date largest-sample self-paced reading study that \textcolor{blue}{aimed to investigate} the use of syntactic and semantic features during subject-verb dependency formation. \textcolor{blue}{However, reading times started to differ before this critical dependency could be formed. Starting at the distractor which intervened between subject and verb, high semantic interference conditions (animate distractors) were read slower than low semantic interference conditions (inanimate distractors). Because the distractor differed between conditions, we did not analyze distractor reading times. Instead, we treat the pre-critical region which was identical across conditions and preceded the critical verb as a proxy to investigate the difference starting at the distractor. Bayes factors provided extremely strong evidence for a semantic interference effect in the pre-critical region. In contrast, there was evidence against syntactic interference and an interaction in that region. We refrain from interpreting effects which were observed in later regions, i.e., the critical verb and spill-over region, because it is unclear whether they were caused by the difference starting at the distractor or other processes. The reading times differences at the distractor and the pre-critical region are discussed below.}}

@@

Indeed, previous work using the same design as the present paper had also found semantic interference effects at the pre-critical region: both \cite{vandyke07} and \cite{mertzen} found such effects. \citeauthor{mertzen}'s (\citeyear{mertzen}) Figures 5 and 6 indicated reading time differences in earlier regions, especially for their German data (see their Figure 6). \cite{vandyke07} did not report reading times for the whole sentences / distractors;  \citeauthor{vandyke07} attributed the effects observed at the pre-critical region to plausibility differences between conditions, but as \cite{mertzen} also pointed out, encoding interference could be an explanation even in that study. Given these earlier findings, our results are consistent with the encoding interference explanation.

@@

\section*{Conclusion}
This project investigated the use of  syntactic and semantic cues during subject-verb dependency resolution with self-paced reading and event-related potentials. Both our experiments consistently showed semantic interference, i.e., more processing difficulty in the presence of a semantically matching distractor. This manifested in longer self-paced reading times starting at the distractor \textcolor{blue}{itself which persisted almost until the end of the sentence} and in a more negative N400 amplitude for the high vs.\ low semantic interference conditions at the critical verb. The observed differences in the time course of the effect are likely due to the methods used in the two sets of experiments (self-paced vs.\ auto-paced reading), which might have lead to a different mixture of encoding and retrieval interference. \Copy{thisdesign3}{Overall, \textcolor{blue}{in the present design}, we found no decisive evidence for the use of a syntactic cue such as \{$\pm$ grammatical subject\}, as was previously assumed in the literature; computational modeling shows that -- at least given the present data -- the parser uses the syntactic cue  \{$\pm$ subject-in-same-clause\}  to identify the correct target for retrieval}\label{thisdesign3}.  A broader implication is that the parser may be able to use hierarchical structure in the input to target the correct syntactic dependent from memory, leading to no syntactic interference from a distractor that occurred within an embedded clause.

>>>>

\textcolor{blue}{Crucially, the reading times difference starting at the distractor} persisted in the following regions. The distractor and the immediately following regions differed between conditions, \textcolor{blue}{therefore we did not analyze their reading times. In our materials, there were two words between the distractor and the critical verb which were identical across conditions: the pre-pre-critical and pre-critical adverbs. The pre-pre-critical region was introduced in the present study to absorb potential effects caused by processing the clause boundary of the embedded clause, rendering it a problematic region for analyses. The pre-critical region was not adjacent to clause boundaries. The reading times of the pre-critical region were analyzed to investigate the reading times difference starting at the distractor. This difference starting at the distractor and persisting almost until the end of the sentence was problematic for the analysis and interpretation of reading times in the critical (and spill-over) region. Any effects that might be present there cannot be attributed clearly to the processing of the respective region. However, for comparability with previous work, we briefly also report analyses of the reading times of the critical and spill-over region.}

@@

\paragraph{Hypothesis testing using Bayes factors}
\textcolor{blue}{To complete the analysis of the effects in the pre-critical, critical and spill-over regions -- although they were confounded by the long-lasting reading time differences starting at the distractor, we here present a Bayes factor analysis using all the data from Experiments 1a and 1b (see Figure \ref{fig:spr_bfs}). There was} very strong to extreme evidence for the semantic interference effect in the pre-critical and spill-over region (BF$_{10}$ between 36 and 231). In the critical region, the evidence for semantic interference was weak \textcolor{blue}{at best} (Normal(0, 0.01): BF$_{10}$ = 3.5, Normal(0, 0.05): BF$_{10}$ = 1.3, Normal(0, 0.1): BF$_{10}$ = 0.7). There was evidence against an effect of syntactic interference under almost all priors in all regions (BF$_{10}$ $<$ 0.5). Only in the spill-over region when small effects between -8 and 8 ms were assumed a priori, there was anecdotal evidence for syntactic interference (BF$_{10}$ = 2.8). There was overall evidence against the interaction of syntactic and semantic interference (BF$_{10}$ $<$ 1). Under the narrowest prior in the critical region, the Bayes factor provided no evidence either way (neither for nor against the interaction (BF$_{10}$ = 1.4)).

@@

\subsection*{Discussion}
\Copy{SPR_discussion_revised}{We presented the to-date largest-sample self-paced reading study that \textcolor{blue}{aimed to investigate} the use of syntactic and semantic features during subject-verb dependency formation. \textcolor{blue}{However, reading times started to differ before this critical dependency could be formed. Starting at the distractor which intervened between subject and verb, high semantic interference conditions (animate distractors) were read slower than low semantic interference conditions (inanimate distractors). Because the distractor differed between conditions, we did not analyze distractor reading times. Instead, we treat the pre-critical region which was identical across conditions and preceded the critical verb as a proxy to investigate the difference starting at the distractor. Bayes factors provided extremely strong evidence for a semantic interference effect in the pre-critical region. In contrast, there was evidence against syntactic interference and an interaction in that region. We refrain from interpreting effects in the critical and spill-over region, because it is unclear whether they were caused by the difference starting at the distractor or other processes, i.e., memory retrieval. The reading times differences at the distractor and the pre-critical region are discussed below.}}

@@

\Copy{previous_encoding}{Indeed, previous work using the same design as the present paper had also found semantic interference effects at the pre-critical region: both \cite{vandyke07} and \cite{mertzen} found such effects. \citeauthor{mertzen}'s (\citeyear{mertzen}) Figures 5 and 6 indicated reading time differences in earlier regions, especially for their German data (see their Figure 6). \cite{vandyke07} did not report reading times for the whole sentences / distractors;  \citeauthor{vandyke07} attributed the effects observed at the pre-critical region to plausibility differences between conditions, but as \cite{mertzen} also pointed out, encoding interference could be an explanation even in that study. Given these earlier findings, our results are consistent with the encoding interference explanation.}\label{previous_encoding}

@@

\section*{Conclusion}
This project investigated the use of  syntactic and semantic cues during subject-verb dependency resolution with self-paced reading and event-related potentials. Both our experiments consistently showed semantic interference, i.e., more processing difficulty in the presence of a semantically matching distractor. This manifested in longer self-paced reading times starting at the distractor \textcolor{blue}{itself which persisted almost until the end of the sentence} and in a more negative N400 amplitude for the high vs.\ low semantic interference conditions at the critical verb. The observed differences in the time course of the effect are likely due to the methods used in the two sets of experiments (self-paced vs.\ auto-paced reading), which might have lead to a different mixture of encoding and retrieval interference. \Copy{thisdesign3}{Overall, \textcolor{blue}{in the present design}, we found no decisive evidence for the use of a syntactic cue such as \{$\pm$ grammatical subject\}, as was previously assumed in the literature; computational modeling shows that -- at least given the present data -- the parser uses the syntactic cue  \{$\pm$ subject-in-same-clause\}  to identify the correct target for retrieval.}\label{thisdesign3}  A broader implication is that the parser may be able to use hierarchical structure in the input to target the correct syntactic dependent from memory, leading to no syntactic interference from a distractor that occurred within an embedded clause.
